---
title: 'Week 19: Genome Informatics'
author: 'Stefanie Hodapp (PID: A53300084)'
date: "12/6/2021"
output:
  pdf_document: default
  html_document: default
---

## Population Scale Analysis

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
# read this .txt file into R
df <- read.table("rs8067378_ENSG00000172057.6.txt")
head(df)

# determine the sample size for each genotype
table(df$geno)        # A/A: 108, A/G: 233, G/G: 121

# determine corresponding median expression levels for each of these genotypes
library(dplyr)

# filter data for each genotype
df_AA <- filter(df, geno == "A/A")     
head(df_AA)

df_AG <- filter(df, geno == "A/G")     
head(df_AG)

df_GG <- filter(df, geno == "G/G")     
head(df_GG)

# Calculate median expression levels for each genotype
summary(df_AA)                         
median(df_AA$exp)         # 31.25

summary(df_AG)                 
median(df_AG$exp)         # 25.065

summary(df_GG)                  
median(df_GG$exp)         # 20.074
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

> From this boxplot, you can infer that the genotype affects ORMDL3 expression, in which G/G corresponds with the least expression while A/A corresponds with the greatest expression. Based on these data, this SNP does effect the expression of ORMDL3.  

```{r}
boxplot(exp~geno, data=df, xlab="Genotype", ylab="Expression", col=2:4, notch = TRUE)

stripchart(exp~geno, data=df, method = "jitter", pch = 16, vertical = TRUE, add = TRUE)
```





