---
title: "class10"
author: 'Stefanie Hodapp (PID: A53300084)'
date: "10/29/2021"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
  '': default
---

# 1. Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
>Q1. How many different candy types are in this dataset?
>Q2. How many fruity candy types are in the dataset?

```{r}
nrow(candy) # Q1. There are 85 different candy types in this dataset

table(candy$fruity) # Q2. There are 38 fruity candy types
```

# 2. What is your favorite candy?

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
>Q4. What is the winpercent value for “Kit Kat”?
>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["ReeseÕs Peanut Butter cup", ]$winpercent # Q3. 84.18029%
candy["Kit Kat", ]$winpercent # Q4. 76.7686%
candy["Tootsie Roll Snack Bars", ]$winpercent # Q5. 49.6535%
```

Use the skim() function in the skimr package to give a quick overview of the candy dataset.

```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
> There are 3 variables that are on a different scale to the others: sugarpercent, pricepercent, and winpercent. These variables are on a 0 to 1 scale representing percentages, whereas whereas the other variables are either 0 or 1 values.

>Q7. What do you think a zero and one represent for the candy$chocolate column?
>A zero value represents that the candy does not contain chocolate, while a one value represents that the candy bar contains chocolate.  

>Q8. Plot a histogram of winpercent values
>Q9. Is the distribution of winpercent values symmetrical?

```{r}
hist(candy$winpercent) # Plotting a histogram using hist()

library(ggplot2)
ggplot(candy) + aes(x=winpercent) + geom_histogram() 
# Plotting a histogram using ggplot()

# Q9. The distribution of the winpercent values is asymmetrical (it is slightly skewed to the right)
```

>Q10. Is the center of the distribution above or below 50%?
>The center of the distribution is below 50%.

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# Using the mean function to compare chocolate and fruity win percentages

mean_chocolate <- mean(candy$winpercent[as.logical(candy$chocolate)])
print(mean_chocolate)

mean_fruity <- mean(candy$winpercent[as.logical(candy$fruity)])
print(mean_fruity)

mean_chocolate > mean_fruity # Chocolate candy is higher ranked than fruit candy (60.92153 vs 44.11974)
```
>Q12. Is this difference statistically significant?

```{r}
# Using the T test function to compare chocolate and fruity win percentages 

x <- candy$winpercent[as.logical(candy$chocolate)]
y <- candy$winpercent[as.logical(candy$fruity)]
t.test(x,y) # This difference is statistically significant (p-value = 2.871e-08)
```

# 3. Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?
>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent),], n=5) # 5 least liked candy types: Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters
head(candy[order(candy$winpercent, decreasing = TRUE),], n=5) # 5 all time favorite candy types: Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, Snickers. 
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col() + labs(x = "win percent", y = "candy")
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col() + labs(x = "win percent", y = "candy")
```

Setup a color vector to be used in future plots. Start by making a vector of all black values (one for each candy). Then overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

Fill the previous bar plot with these colors

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

>Q17. What is the worst ranked chocolate candy?
>Sixlets

>Q18. What is the best ranked fruity candy?
>Starbursts

# 4. Taking a look at pricepercent

Plot of winpercent vs the pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
 ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord_winpercent <- order(candy$winpercent, decreasing = TRUE)
head(candy[ord_winpercent,c(11,12)], n=10)

# Reese's Miniatures have the second highest winpercent (81.86626) and a pricepercent of 0.279.
```

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord_pricepercent <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord_pricepercent,c(11,12)], n=5)

# The top 5 most expensive candies are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. The least popular of these is Nik L Nip. 
```

>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy),pricepercent)) +
  geom_col() + labs(x = "price percent", y = "candy")
```
```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point() + labs(x="price percent", y="candy")
```

# 5 Exploring the correlation structure

Load corrplot package

```{r}
library(corrplot)
```

Plot a correlation matrix using the candy dataset

```{r}
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
>Two variable that are antivorrelated are chocolate and fruity

>Q23. Similarly, what two variables are most positively correlated?
>Two variables that are most positively correlated are chocolate and bar. 

# 6. Principal Component Analysis

Apply PCA using the prcomp() function to the candy dataset

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Plot PC1 vs PC2

```{r}
plot(pca$x[,1:2])
```

Change the plotting character and add some color using the color vector define previously.

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

Make a new data-frame with the PCA results and candy data to be used with the ggplot() function

```{r}
my_data <- cbind(candy, pca$x[,1:3])
head(my_data)
```

Use ggplot() to make a plot with our PCA results

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

Use the ggrepel package and the function ggrepel::geom_text_repel() to label the plot with non overlapping candy names, a title, and subtitle.

```{r}
library(ggrepel)

 p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
 labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")

```

Pass the ggplot object p to plotly to generate an interactive plot that you can mouse over to see labels

```{r}
library(plotly)
# ggplotly(p)
```

Look at PCA our loadings. 

```{r}
par(mar=c(8,4,2,2)) # Set the margins of the graph by calling the par() function with the mar argument. 
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution") 
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
>The variables that are picked up stringly by PC1 in the positive direction are fruity, hard, and pluribus. These make sense since many hard candies are fruit flavored and are sold as multiples in one package (e.g. jolly ranchers). 

