---
title: "Class17 Mini Project"
author: 'Stefanie Hodapp (PID: A53300084)'
date: "12/3/2021"
output:
  pdf_document: 
    latex_engine: xelatex
---

# 1. Background

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

# 2. Getting Started

> Q1. What column details the total number of people fully vaccinated?

```{r}
which(colnames(vax)=="persons_fully_vaccinated") # Column 9
```

> Q2. What column details the Zip code tabulation area? 

```{r}
which(colnames(vax)=="zip_code_tabulation_area" ) # Column 2
```

> Q3. What is the earliest date in this dataset?

```{r}
min(vax$as_of_date) # 2021-01-05
```

> Q4. What is the latest date in this dataset?

```{r}
max(vax$as_of_date) # 2021-11-30
```

> Q5. How many numeric columns are in this dataset? 

```{r}
# call the skim() function from the skimr package to get a quick overview of this dataset
skimr::skim(vax)
```

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

```{r}
sum(is.na(vax$persons_fully_vaccinated)) # 8472
```

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
signif((sum(is.na(vax$persons_fully_vaccinated)) / nrow(vax))*100, 2) # 10%
```
# 2.1 Working with Dates 

```{r}
library(lubridate)
```

```{r}
today()
```

```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)

# Compute the number of days that have passed since the first vaccination reported in this dataset
today() - vax$as_of_date[1]

# Determine how many days the dataset span
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset? 
> Time difference of 332 days

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 
> Time difference of 329 days

# 3. Working with ZIP codes

```{r}
# library(zipcodeR)
```

```{r}
# find the centroid of the La Jolla 92037 (i.e. UC San Diego) ZIP code area
#geocode_zip('92037')

# Calculate the distance between the centroids of any two ZIP codes in miles, e.g.
# zip_distance('92037','92109')

# pull census data about ZIP code areas 
# reverse_zipcode(c('92037', "92109") )
```

# 3.1 Focus on the San Diego area

```{r}
# Subset to San Diego county only areas
sd <- vax[ vax$county == "San Diego" , ]
nrow(sd)
```

```{r}
# subset all San Diego county areas with a population of over 10,000
library(dplyr)
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area)) # 107 unique zip codes
```

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
sd$zip_code_tabulation_area[which.max(sd$age12_plus_population)] # 92154
```

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”?

```{r}
# Filter data by as_of_date == 2021-11-09
sd_nov_9 <- filter(sd, as_of_date == "2021-11-09")
head(sd_nov_9)

# Calculate the average of percent of population fully vaccinated for all of san diego county
mean(sd_nov_9$percent_of_population_fully_vaccinated, na.rm=TRUE)*100 # 67.40809%
```

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?

```{r}
# histogram using base R
hist(sd_nov_9$percent_of_population_fully_vaccinated, main ="San Diego County Vaccination Rates", xlab ="% Fully Vaccinated as of Nov 9, 2021", ylab = "Count")
```

```{r}
# histogram using ggplot
# ggplot(data=sd_nov_9, aes(x=percent_of_population_fully_vaccinated)) + geom_histogram() + labs(title="San Diego County Vaccination Rates", x="% Fully Vaccinated as of Nov 9, 2021", y = "Count")
```

# 3.1.1 Focus on UCSD/La Jolla

```{r}
# Filter data for UC San Diego (UCSD resides in the 92037 ZIP code area and is listed with an age 5+ population size of 36,144)
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
head(ucsd)
```
> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
library(ggplot2)

ggplot(ucsd) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x= "Date", y="Percent Vaccinated")
```

# 3.1.2 Comparing to similar sized areas

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")

head(vax.36)
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
# Calculate mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”.
mean_pop <- mean(vax.36$percent_of_population_fully_vaccinated)
mean_pop

# Add mean_pop as a straight horizontal line to your plot from above with the geom_hline() function
ggplot(ucsd) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x= "Date", y="Percent Vaccinated") +
  geom_hline(yintercept=mean_pop, linetype = 'dashed', col = 'red')
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) +
  aes(x=percent_of_population_fully_vaccinated) +
  geom_histogram(bins=20) +
  labs(x= "Percent Vaccinated", y="Count")

```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?
> The 92040 zip code is below the average value (0.52142 < 0.6645132) while the 92109 zip code is above the average value (0.68912 > 0.6645132). 

```{r}
# 92040
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)   # 0.52142

# 92109
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)   # 0.68912
```

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination rate across California",
       subtitle="Only areas with a population above 36k shown") +
  geom_hline(yintercept = mean_pop, linetype="dashed")
```

> Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?
> N/A, but I am glad we had our last class in-person! Thank you for an excellent quarter! 

