---
title: 'Class 6: R Functions'
author: 'Stefanie Hodapp (PID: A53300084)'
date: "10/15/2021"
output:
  pdf_document: default
  html_document: default
---

## Quick Rmarkdown intro

We can write text of course just like any file. We can **style text to be bold** or *italic*. 

Do :

- this
- and that
- and another thing

This is more text  
and this is a new line

---

We can include some code:

```{r}
plot(1:10)
```


## Time to write a function

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

First I want to find the lowest score. I can use the **min()** to find it and the **which.min()** function to find where it is (i.e. its position in the vector)
```{r}
which.min(student1)
```

```{r}
student1[ -which.min(student1) ]
```

Now I can call the **mean()** function to get the average.

```{r}
mean(student1[ -which.min(student1) ])
```

Does this work for student2?
```{r}
mean(student2[ -which.min(student2) ], na.rm=TRUE)
```

One great idea is to replace the na values with zero. Let's do it.
```{r}
which(is.na(student2))
```

This is.na() function returns a logical vector where TRUE elements indicate thepresence of NA values. 
```{r}
is.na(student2)
```

Let's replace NAs with 0 and print the new values in this
```{r}
student2_new <- student2
student2_new[is.na(student2_new)] = 0
student2_new
mean(student2_new[ -which.min(student2_new)])
```

How about student 3?
```{r}
student3_new <- student3
student3_new[is.na(student3_new)] = 0
student3_new
mean(student3_new[ -which.min(student3_new)])
```

We can make the object names more clear

```{r}
x <- student3
x[is.na(x)] = 0
mean(x[ -which.min(x)])
```

A vector containing an accidental character can be changed to a numeric vector using the following function:
```{r}
student4 <- c(100, NA, 90, "90", 90, 90, 97, 80)
```

```{r}
x <- student4
x <- as.numeric(x)
x[is.na(x)] = 0
mean(x[ -which.min(x)])
```

Now we can write our function: 
All functions have at least 3 things.
A name, an input arguments, and a body.
```{r}
grade <- function(x) {
  x <- as.numeric(x)
  x[ is.na(x)] = 0
  mean(x[ -which.min(x)])
}
```

```{r}
grade(student1)
```

Develop a function for calculating average grades for fictional students in a fictional class.

Inpur CSV file containing the class gradebook
```{r}
scores <- read.csv("https://tinyurl.com/gradeinput", row.names=1)
scores
```

Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score.

We are going to use the super useful **apply()** function to grade all the students with our **grade()** function.

Question 1
```{r}
Q1 <- apply(scores, 1, grade)
Q1
```
Question 2
```{r}
Q2 <- which.max(Q1)
Q2
```

Q3. Replace or mask Na values to zero
```{r}
mask <- scores 
mask[is.na(mask)] = 0
mask
```

Question 3. Ignore the NA missing values with na.rm=TRUE
```{r}
Q3 <- apply(mask, 2, mean, na.rm=TRUE)
Q3
```

Question 4. From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)?

Here we will use the **cor()** function.
```{r}
cor(mask$hw1, Q1)
```

I can call the **cor()** for every homework and get a value for each but that sucks. Let's use **aply()** and do them all in one go.
```{r}
apply(mask,2,cor, Q1)
```

```{r}
boxplot(scores)
```

Another way to write the grade function... 

```{r}
grade2 <- function(x, drop.lowest=TRUE) {
  x <- as.numeric(x)
  x[ is.na(x) ] = 0
  if (drop.lowest) {
    mean(x[ -which.min(x) ])
  } else {
    mean(x)
  }
}
```

```{r}
student <- c(100, 100, 100, 90)
grade2(student, drop.lowest = FALSE)
```

